Fast Sorting on Modern Computers

### 2005-07-29 docs/tech_report/conclusions.tex
nearly finished

\subsection{Order N squared sorts}
\subsection{Heapsorts}
\subsection{Mergesorts}
\subsection{Quicksorts}
\subsection{Radixsorts}
\subsection{Shellsorts}

\section{Comparison with LaMarca's results}

@@

\section{Instrucyion count results}

>>>>

\subsection{Elementary Sorts}

We found that insertion sort has the best performance of any of the elementary
sorts, with fewer instructions, cache misses and exactly one branch
misprediction per key. Selection sort also has very few branch mispredictions,
and our results match Knuth's calculations. We find that reducing the number of
runs in bubblesort and shakersort at the cost of additional instructions is
counter productive, and that their branch prediction performance is incredibly
poor.

\subsection{Heapsorts}

The cache optimized heaps perform better than the non-optimized versions. At the
cost of a slightly increased instruction count, the number of branch misses can
be greatly reduced, and the overall performance of our 8-heap is significantly
better than the unoptimized version. This is despite our finding of that
increasing the fanout increases branch mispredictions.

Additionally, we created an improvement to heapsort which removed the need for a
sentinel, without introducing any more instructions.

\subsection{Mergesorts}

We found that algorithm S has a lower instruction count than algorithm N, but
that this does not necessarily translate to better performance. Our results show
also that the level 1 cache needs to be taken into account, that the extra
instruction in multi-mergesort make it slower than double tiled mergesort, and
that multi-mergesort has regular branch predictive access patterns during the
k-way merge, which can be taken advantage of by a 2-level adaptive predictor.

\subsection{Quicksorts}

Our results show base quicksort being the fastest of the quicksorts. We note our
lowest instruction count occurs with the median-of-7 partitioning. We also see
that using a binary search slows multi-quicksort down compared to a sequential
search, but that its extra instructions make it slower than tuned quicksort.
Despite being slower in our tests, pur simulations show tuned quicksort to have
fewer cache misses, instructions and branch mispredictions than base quicksort.

\subsection{Radixsorts}

We were able to reduce the number of cache misses considerably using a technique
adapted from LaMarca's tiled mergesort, and managed to make cache radixsort the
fastest sort in our tests. This is due to its low instruction count, few cache
misses and lack of branch mispredictions. Although we attempted to align it
using another technique from mergesort, we found this decreased its performance.

\subsection{Shellsorts}

Based on our tests, shellsort performs very similarly to an $O(NlogN)$ sort,
outperforming heapsort. We note that our improved version is considerably
faster, mostly due to its reduced instruction count.

We were also able to measure it's branch mispredictions, and determine a formula
to calculate them.

\section {Branch Results}

No results for the branch prediction performance of sorting algorithm yet
exists. We were able to find some interesting results.

Insertion sort has exaclty one miss per key. Using this result we were able to
make several other sorts faster by replacing their simple inline sorts with
insertion sorts.

Selection sort also has a very low number of branch mispredictions, the results
of which match up well to Knuth's calculation. We found aswell that bubblesort
and shakersort have incredibly bad branch prediction performance, bring
$O(N^2)$.

We found a pattern in branch misprediction when using a fanout in our 8-heap,
and that this pattern matched closely that of selection sort, discussed above.

TODO merge

TODO quick

TODO radix

TODO shell

%TODO results from software predictors

\section{Comparison with LaMarca's Results}

@@

\section{Instruction count results}
